---
title: 'Comparing the Amount of Water Needed to Produce Different Foods '
author: 'Sancho '
date: '2020-03-01'
slug: water-needed-to-produce-food
categories: []
tags:
  - Food
  - Environmental
subtitle: ''
summary: ''
authors: []
lastmod: '2020-03-01T19:49:32-05:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
output:
  blogdown::html_page:
    toc: yes
draft: true
---

```{r include=FALSE, eval=FALSE}
# render magic (copied from ellocke, not exactly sure what this does)
rmarkdown::render("index.Rmd")
```

```{r}
library(tidyverse)
library(maps)

#links
#https://www.usgs.gov/special-topic/water-science-school/science/how-much-water-there-earth?qt-science_center_objects=0#qt-science_center_objects
#https://www.usgs.gov/special-topic/water-science-school/science/where-earths-water?qt-science_center_objects=0#qt-science_center_objects
#https://waterfootprint.org/en/resources/waterstat/product-water-footprint-statistics/


```

#Why should we care about water? 

I remember my first reaction to hearing about water scarcity was doubt. My first through was: isn't 70% of the whole planet covered in water? How can we possibly have a problem. Then, in January of 2018 it was announced that Cape Town, South Africa was less than 90 days away from running out of water, which they called "Day Zero" [(Time, 2018)](https://time.com/5103259/cape-town-water-crisis/). At that point they would have to shut off water and people would have to go to water collections points, then wait in line to receive their daily allotment of 6.6 gallons. For reference, shower heads in the US are usually capped at 2.5 gallons a minute [(Portland Water Bureau)](https://www.portlandoregon.gov/water/article/305153), which means a 2 minute and 38 second shower would use up all of the allowance. 

##Fresh vs Salt Water


```{r}

#get the world map data
world <- map_data("world")

ggplot(data = world) +
    geom_sf() +
    coord_sf(xlim = c(-102.15, -74.12), ylim = c(7.65, 33.97), expand = FALSE)

33.9249° S, 18.4241° E

```

Fortunately, through several measures Capetonians were able to redirect water and reduce their use enough to push back Day Zero into the rainy season. When the rains could refill the reservoir and they were able to postpone Day Zero indefinitely. One measure taken was every person was capped at 23 gallons per day, which is about 9min and 12 seconds of shower time per person. Businesses also helped with the effort, restaurants put up "if its yellow let it mellow" signs, and offices had "dirty shirt" challenges to see who could wear the same shirt the longest without washing [(Alexander, 2019)](https://www.citylab.com/environment/2019/04/cape-town-water-conservation-south-africa-drought/587011/). However, one single change by the largest consumers of water was able to reduce water usage by 15%. Farmers agreed to half the water they were using from 30% of all water down to 15% [(Mail and Guardian, 2018)](https://mg.co.za/article/2018-02-05-day-zero-pushed-back-to-may-11/). 

#Water Use in Agriculture



```{r}
library(ggplot2)
library(dplyr)
library(maps)


```
